---
title: "OpenSAFELY COVID-19 Vaccine Coverage Snapshot"
output: html_document
---

<br>

OpenSAFELY is a secure analytics platform for electronic patient records built on behalf of NHS England to deliver urgent academic and operational research during the pandemic.

This snapshot of vaccine coverage is an extension of our [previous weekly reports](https://reports.opensafely.org/reports/vaccine-coverage-thirdbooster-doses/) on COVID-19 vaccination uptake in England. The snapshot includes data from the 40% of general practices that use TPP electronic health record software. The data requires careful interpretation and there are a number of caveats. Please read the full detail about our methods and discussion of our earlier results in our [peer-reviewed publication in the British Journal of General Practice](https://bjgp.org/content/72/714/e51).

The full analytic methods behind the latest results in this report are available [here](https://github.com/opensafely/vaccine-counts).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# Import libraries
library('here')
library('tidyverse')
library('gt')
library('gtsummary')
library('cowplot')
library('RColorBrewer')
library('ggridges')

# Function fct_case_when 
source(here("analysis", "snapshot_utils", "fct_case_when.R"))

# Read in full data for density plots
d_full <- read_rds(here::here("output", "snapshot", "processed_snapshot.rds")) %>%
  # Define age_plot for facets
  mutate(
    age_plot =  fct_case_when(
        agegroup_medium=="75+" ~ "75y+",
        agegroup_medium=="65-74" ~ "65-74y",
        agegroup_medium=="50-64" ~ "50-64y",
        agegroup_medium=="18-49" ~ "18-49y",
        TRUE ~ "0"
      )
  )

# Clean time since vaccination variable
d_full$time_since_last_vax_months[d_full$time_since_last_vax_months>24] = 30

# Read in summary data and define clean variables for tabulations
d <- read_rds(here::here("output", "snapshot_report", "snapshot_summary.rds")) %>%
  mutate(Clean_N = formatC(N, big.mark=",", format="f", digits=0),
         Clean_0 = formatC(Dose_0, big.mark=",", format="f", digits=0),
         Clean_1 = formatC(Dose_1, big.mark=",", format="f", digits=0),
         Clean_2 = formatC(Dose_2, big.mark=",", format="f", digits=0),
         Clean_3 = formatC(Dose_3, big.mark=",", format="f", digits=0),
         Clean_4 = formatC(Dose_4, big.mark=",", format="f", digits=0),
         Clean_5plus = formatC(Dose_5plus, big.mark=",", format="f", digits=0),
         Clean_past_12m = formatC(Vax_past_12m, big.mark=",", format="f", digits=0),
         Clean_past_24m = formatC(Vax_past_24m, big.mark=",", format="f", digits=0),
         Perc_0 = round(Dose_0/N*100,1),
         Perc_1 = round(Dose_1/N*100,1),
         Perc_2 = round(Dose_2/N*100,1), 
         Perc_3 = round(Dose_3/N*100,1), 
         Perc_4 = round(Dose_4/N*100,1), 
         Perc_5plus = round(Dose_5plus/N*100,1), 
         Perc_past_12m = round(Vax_past_12m/N*100,1), 
         Perc_past_24m = round(Vax_past_24m/N*100,1),
         `0` = paste0(Clean_0," (",Perc_0,"%)"),
         `1` = paste0(Clean_1," (",Perc_1,"%)"),
         `2` = paste0(Clean_2," (",Perc_2,"%)"),
         `3` = paste0(Clean_3," (",Perc_3,"%)"),
         `4` = paste0(Clean_4," (",Perc_4,"%)"),
         `5+` = paste0(Clean_5plus," (",Perc_5plus,"%)"),
         `Dose count, median (IQR)` = paste0(Median_dose_count," (",
                                             Q1_dose_count,"-",Q3_dose_count,")"),
         `Vaccinated in past 12 months, n (%)` = paste0(Clean_past_12m," (",Perc_past_12m,"%)"),
         `Vaccinated in past 24 months, n (%)` = paste0(Clean_past_24m," (",Perc_past_24m,"%)"),
         `Time in months since last dose, median (IQR)` = paste0(Median_time_since_last_dose," (",
                                                              Q1_dose_time_since_last_dose,"-", 
         Q3_dose_time_since_last_dose,")")
  ) %>%
  rename(Integer_N = N, N = Clean_N)

# Clean redactions
d[d=="NA (NA%)"] = "[Redacted]"
d[d=="NA (NA-NA)"] = "[Redacted]"

## Set factor levels
d$Covariate = factor(d$Covariate, levels=unique(d$Covariate))
d$Level = factor(d$Level, levels=unique(d$Level))
d$Population = factor(d$Population, levels=rev(unique(d$Population)))

# Create data subsets
d_all <- subset(d, Population=="All")
d_18to49 <- subset(d, Population=="18-49")
d_50to64 <- subset(d, Population=="50-64")
d_65to74 <- subset(d, Population=="65-74")
d_75plus <- subset(d, Population=="75+")

# Create long-form table
d_long = rbind(
  d %>% mutate(Count = "0", Perc = Perc_0),
  d %>% mutate(Count = "1", Perc = Perc_1),
  d %>% mutate(Count = "2", Perc = Perc_2),
  d %>% mutate(Count = "3", Perc = Perc_3),
  d %>% mutate(Count = "4", Perc = Perc_4),
  d %>% mutate(Count = "5+", Perc = Perc_5plus)
)

# Set factor levels for plotting
d_long$Count = factor(d_long$Count, levels=c("5+", "4", "3", "2", "1", "0"))

# Create colour palette
pal = brewer.pal(n = 10, name = "RdBu")
```

<br><br>

#### Coverage snapshot - 01 September 2023

-   This report summarises the vaccination history of adults in England as of 01 September 2023 - immediately prior to the Autumn 2023 COVID-19 vaccine booster campaign. A summary of COVID-19 vaccine implementation for adults in the UK is given at the bottom of this page. 

-   Age is calculated at 01 September 2023. Age groups (18-49, 50-64, 65-74, and 75+) reflect the eligibility cut-offs used over successive booster campaigns in the UK, as described below.

-   Clinical subgroups are defined using codelists described in our [peer-reviewed publication in the Lancet Public Health](https://www.thelancet.com/journals/lanpub/article/PIIS2468-2667(23)00079-8/fulltext).

-   Index of multiple deprivation is a measure of relative deprivation based on the local area in which a patient lives.

-   We do not distinguish between vaccine products or manufacturers (Oxford/AstraZeneca, Pfizer/BioNTech, or Moderna).

-   For calculations of median time since last dose, unvaccinated individuals were assigned a value of 36 months (exceeding the maximum possible value among vaccinated individuals).

-   Patient counts are rounded to the nearest 7. Rounded counts of <=7 are redacted.

<br><br>

#### Vaccination history by age group {.tabset}

Click on the tabs to display the data as a figure or table. Abbreviations: IQR, interquartile range; Unvax, unvaccinated.

##### Summary figure

*In the plot on the right, the height of the lines represents the relative proportion of individuals.*

```{r, fig.width=12, fig.height=3.5}
# Function with basic plot parameters for stacked count plot
g1_extras = function(gg) {
  gg + 
    geom_bar(stat="identity", position="stack") +
    coord_flip() +
    theme_bw() + 
    scale_fill_manual(values=pal[c(10,9,8,3,2,1)]) +
    ggtitle("Dose count") + ylab("%") + xlab("Age (y)") +
    theme(title = element_text(size = 10),
          legend.position = "bottom",
          legend.title = element_blank(),
          legend.text = element_text(size = 13),
          strip.background = element_blank(),
          strip.text = element_blank(),
          panel.spacing = unit(1, "lines"),
          axis.text = element_text(size = 14),
          axis.title = element_text(size = 14), 
          axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0))) +
    guides(fill = guide_legend(nrow = 1, reverse = TRUE))
}

# Function with basic plot parameters for prevalence plot
g2_extras = function(gg) {
  gg + 
    geom_bar(stat="identity", position="dodge", fill=pal[10]) +
    coord_flip() +
    theme_bw() + 
    ylim(0,100) +
    ggtitle("Vaccinated in past 12 months (%)") + ylab("%") + xlab(" ") +
    theme(title = element_text(size = 10),
          strip.background = element_blank(),
          strip.text = element_blank(),
          panel.spacing = unit(1, "lines"),
          axis.text.x = element_text(size = 14),
          axis.text.y = element_blank(),
          axis.title = element_text(size = 14))
}

# Function with basic plot parameters for ridge plot
g3_extras = function(gg) {
  gg + 
    #geom_violin(fill=pal[10], alpha=0.2) +  
    geom_density_ridges(colour=pal[10], fill=pal[10], alpha=0.2, scale=0.5, bandwidth=1) +
    theme_bw() +
    #scale_x_reverse(limits=c(30,0), breaks=c(30,24,18,12,6,0), labels=c("Unvax/\n>24","24","18","12","6","0")) +
    scale_x_continuous(limits=c(0,30), breaks=c(0,6,12,18,24,30), labels=c("0","6","12","18","24","Unvax/\n>24")) +
    ggtitle("Time since last dose") + ylab(" ") + xlab("Months") +
    theme(title = element_text(size = 10),
          strip.background = element_blank(),
          strip.text = element_text(size = 14),
          panel.spacing = unit(1, "lines"),
          axis.text.x = element_text(size = 14),
          axis.text.y = element_blank(),
          axis.title = element_text(size = 14),
          plot.margin = margin(r = 20))
}

# Create plots for vaccination history by age group
g1 = ggplot(subset(d_long, Population=="All"), aes(x=Level, y=Perc, fill = Count)) %>% 
  g1_extras()
g2 = ggplot(d_all, aes(x=Level, y=as.numeric(Perc_past_12m))) %>% 
  g2_extras()
g3 = ggplot(d_full, aes(y=age_plot, x=as.numeric(time_since_last_vax_months))) %>% 
  g3_extras() + scale_y_discrete(limits=rev)
plot_grid(g1, g2, g3, ncol=3, align="h", axis="tb", rel_widths=c(1.5,1,1.2))
```

<br><br>

##### Summary table

```{r}
d_all %>% 
  rename(`Age group` = Level) %>%
  arrange(rev(`Age group`)) %>%
  select(`Age group`, N, `0`, `1`, `2`, `3`, `4`, `5+`, `Dose count, median (IQR)`, 
         `Vaccinated in past 12 months, n (%)`, `Vaccinated in past 24 months, n (%)`, `Time in months since last dose, median (IQR)`) %>%
  gt() %>%
  tab_spanner(
    label = "Dose count",
    columns = c("0", "1", "2", "3", "4", "5+")
  ) %>%
   tab_options(row_group.font.weight="bold", column_labels.font.weigh="bold") %>%
   tab_options(table.font.size = pct(95))
```

<br><br>

#### Vaccination history by age group, ethnicity, and deprivation {.tabset}

Click on the tabs to display summary figures and tables. The figures provide a summary of data broken down by ethnicity and deprivation index. The tables provide a more detailed breakdown of vaccination history according to demographic and clinical subgroups (including more narrowly defined ethnic groups). Abbreviations: CKD, chronic kidney disease; IMD, index of multiple deprivation; IQR, interquartile range; RRT, renal replacement therapy; Unvax, unvaccinated.

##### Figure (ethnicity)

*See tables for vaccination history according to more narrowly defined ethnic groups. In the plots on the right, the height of the lines represents the relative proportion of individuals.*

```{r, fig.width=12, fig.height=10}
g1 = ggplot(subset(d_long, Covariate=="Ethnicity (broad categories)"), 
            aes(x=Level, y=Perc, fill = Count)) %>%
  g1_extras() +
  scale_x_discrete(limits=rev) + xlab("Ethnicity") +
  facet_grid(Population~.,scales = "free_y", space = "free_y") +
  theme(axis.text.y = element_text(size=12))

g2 = ggplot(subset(d, Covariate=="Ethnicity (broad categories)"), 
                   aes(x=Level, y=as.numeric(Perc_past_12m))) %>%
  g2_extras() +
  scale_x_discrete(limits=rev) + 
  facet_grid(Population~.,scales = "free_y", space = "free_y") 

g3 = ggplot(d_full, aes(y=ethnicity, x=as.numeric(time_since_last_vax_months)))  %>%
  g3_extras() +
  scale_y_discrete(limits=rev) +
  facet_grid(age_plot~.,scales = "free_y", space = "free_y")

plot_grid(g1, g2, g3, ncol=3, align="h", axis="tb", rel_widths=c(1.5,1,1.2))
```

<br><br>

##### Figure (deprivation)

```{r, fig.width=12, fig.height=10}
g1 = ggplot(subset(d_long, Covariate=="IMD quintile"), 
            aes(x=Level, y=Perc, fill = Count)) %>%
  g1_extras() +
  scale_x_discrete(limits=rev) + xlab("IMD quintile") +
  facet_grid(Population~.,scales = "free_y", space = "free_y") +
  theme(axis.text.y = element_text(size=12)) 

g2 = ggplot(subset(d, Covariate=="IMD quintile"), 
                   aes(x=Level, y=as.numeric(Perc_past_12m))) %>%
  g2_extras() +
  scale_x_discrete(limits=rev) + 
  facet_grid(Population~.,scales = "free_y", space = "free_y") 

g3 = ggplot(d_full, aes(y=imd, x=as.numeric(time_since_last_vax_months)))  %>%
  g3_extras() +
  scale_y_discrete(limits=rev) +
  facet_grid(age_plot~.,scales = "free_y", space = "free_y")
  
plot_grid(g1, g2, g3, ncol=3, align="h", axis="tb", rel_widths=c(1.5,1,1.2))
```

<br><br>

##### Table (75y+)

```{r}
# Function to summarise vaccination history for age subgroup
subgroup_gt = function(d_subgroup) {
  d_subgroup %>% 
    select(Covariate, Level, N, `0`, `1`, `2`, `3`, `4`, `5+`, `Dose count, median (IQR)`, 
           `Vaccinated in past 12 months, n (%)`, `Vaccinated in past 24 months, n (%)`,
           `Time in months since last dose, median (IQR)`) %>%
    gt(groupname_col = "Covariate") %>%
    tab_spanner(
      label = "Dose count",
      columns = c("0", "1", "2", "3", "4", "5+")
    ) %>%
    tab_options(row_group.font.weight="bold", column_labels.font.weigh="bold") %>%
    tab_options(table.font.size = pct(90))
}

# Generate table
subgroup_gt(d_75plus)
```

<br><br>

##### Table (65-74y)

```{r}
subgroup_gt(d_65to74)
```

<br><br>

##### Table (50-64y)

```{r}
subgroup_gt(d_50to64)
```

<br><br>

##### Table (18-49y)

```{r}
subgroup_gt(d_18to49)
```

<br><br>

#### UK vaccine implementation in adults

An overview of COVID-19 vaccine implementation for UK adults is given below. Additional eligibility criteria can be found in [COVID-19: the green book, chapter 14a](https://www.gov.uk/government/publications/covid-19-the-green-book-chapter-14a).

-   **Primary vaccination (2020-2021):** All adults in the UK were offered a primary 2-dose series, with eligibility starting from December 2020 (over 80s, health and social care workers, and care home residents and staff) and extending to all adults over the early months of 2021.

-   **Extended primary vaccination:** Immunosuppressed individuals were offered a third primary dose from September 2021.

-   **Autumn 2021 booster campaign:** All adults were offered a booster dose, with eligibility starting from September 2021 (over 50s, health and social care workers, and care home residents, among others) and extending to all adults from November 2021.

-   **Spring 2022 booster campaign:** Over 75s, care home residents, and immunosuppressed individuals were offered an additional dose from March 2022.

-   **Autumn 2022 booster campaign:** Over 50s, care home residents and staff, immunosuppressed individuals, people aged 5-49 years in clinical risk groups or in close contact with immunosuppressed individuals, and carers aged 16-49 years were offered an additional dose from September 2022

-   **Spring 2023 booster campaign:** Over 75s, care home residents, and immunosuppressed individuals were offered an additional dose from March 2023.

-   **Autumn 2023 booster campaign:** Over 65s, care home residents, people aged 6 months to 64 years in clinical risk groups, health and social care workers, people aged 12-64 years who are in close contact with immunosuppressed individuals, and carers aged 16-64 years are being offered an additional dose from September 2023.

<br><br>